\textbf{Evaluating RISC-V Enclaves:}\\ Security Without the Slowdown?

### 2025-09-05 draft_presentation.tex
added new diagrams

\begin{frame}{Enclave Overhead and System Behavior}
\small
% First overlay: text only
\only<1>{
\begin{block}{Performance Overhead}
\begin{itemize}
    \item \textbf{Enclave execution} introduces overhead:
        \begin{itemize}
            \item Compute-bound workloads like CoreMark and Dhrystone incur moderate overhead of 1.06$\times$ to 1.61$\times$.
            \item Hybrid workloads (Compute- and memory-bound) like Kyber suffer severe overhead, up to 2.73$\times$.
        \end{itemize}
    %\item \textbf{Memory-bound workloads} experience significant penalties due to enclave-induced memory isolation and security checks.
\end{itemize}
\end{block}
}
% Second overlay: add figure
\only<2>{
\begin{block}{Overhead Breakdown}
\begin{itemize}
    \item Memory isolation via PMP adds fixed latency per memory access.
    \item Context switches between host and enclave domains are costly.
\end{itemize}
\end{block}
}
% \centering
% \textbf{Key Insight:} Overhead varies based on workload characteristics and memory behavior.
% \end{frame}
% Figure appears only on the second slide of this frame
\only<3>{
% \begin{frame}{Performance Trends and Hardware Bottlenecks}
\centering
\begin{figure}[htbp]
\centering
\begin{tikzpicture}
\begin{axis}[
    ybar,
    bar width=14pt,
    width=10cm,
    height=6cm,
    ymin=0, ymax=120,
    ylabel={Relative performance (\%)},
    symbolic x coords={CoreMark,Dhrystone,Kyber},
    xtick=data,
    ymajorgrids,
    grid style=dashed,
    nodes near coords,
    every node near coord/.append style={font=\footnotesize, /pgf/number format/fixed},
    legend style={at={(0.5,-0.15)}, anchor=north, legend columns=-1}
]

% Native: orange, rounded top, tranRecommendationssparency
\addplot[
    fill={rgb,255:red,230; green,159; blue,0},
    fill opacity=0.8,
    draw=black!60,
    rounded corners=2pt
] coordinates {(CoreMark,100) (Dhrystone,100) (Kyber,100)};

% Enclave: blue, rounded top, transparency
\addplot[
    fill={rgb,255:red,0; green,114; blue,178},
    fill opacity=0.8,
    draw=black!60,
    rounded corners=2pt
] coordinates {(CoreMark,95) (Dhrystone,62) (Kyber,37)};

\legend{Native,Enclave}
\end{axis}
\end{tikzpicture}
\caption{Relative performance of enclave vs native execution.}
\end{figure}
}
\end{frame}

\begin{frame}{Impact of Scalability Limitations}
\small
\only<1>{
\begin{block}{CPU Scaling Limits}
\begin{itemize}
    \item Increasing CPU cores \textbf{benefits multi-threaded tasks}, but gains plateau due to context-switch overhead.
    \item For example, \texttt{CoreMark} shows marginal improvements beyond 4 cores.
\end{itemize}
\end{block}
\centering
\begin{tikzpicture}
\begin{axis}[
    ybar,
    bar width=8pt,
    width=0.75\linewidth,
    height=4.2cm,
    enlarge x limits=0.15,
    ymin=95, ymax=125,
    xlabel={CPU cores},
    ylabel={Rel.\ perf.\ (\%)},
    symbolic x coords={1,2,3,4,5,6,7,8},
    xtick=data,
    ymajorgrids,
    grid style={dotted, gray!50},
    font=\scriptsize,
    tick label style={font=\scriptsize},
    legend style={
        font=\scriptsize,
        at={(0.5,-0.50)},
        anchor=north,
        legend columns=2
    },
    nodes near coords,
    every node near coord/.append style={font=\tiny},
    point meta=explicit symbolic
]

@@

\begin{frame}{PMP Entry Exhaustion and Multi-Enclave Scalability}
\small
\only<1>{
\begin{block}{PMP Resource Limitations}
\begin{itemize}
    \item \textbf{Maximum PMP entries}: 16.
    \item Exhausted PMP entries \textbf{prevent new enclave creation}, causing errors even with abundant system resources.
    \item Failure due to PMP exhaustion: \texttt{SBI error codes}.
\end{itemize}
\end{block}
}
\only<2>{
\begin{block}{Scalability Bottleneck}
\begin{itemize}
    \item In \textbf{multi-tenant environments}, PMP exhaustion limits the number of concurrent enclaves.
    \item This creates scalability issues, especially for high-concurrency workloads.
\end{itemize}
\end{block}
}

@@

% \begin{frame}{PMP Entries vs Max Concurrent Enclaves}
% \centering
\only<3>{
\begin{figure}[htbp]
\centering
\begin{tikzpicture}
\begin{axis}[
    width=10cm,
    height=6cm,
    xlabel={PMP Entries},
    ylabel={Concurrent Enclaves},
    xmin=0, xmax=16,
    ymin=0, ymax=16,
    xtick={0,4,8,12,16},
    ymajorgrids,
    grid style=dashed,
    mark options={solid},
    nodes near coords,
    every node near coord/.append style={font=\footnotesize, /pgf/number format/fixed},
    clip=false,
    cycle list={{teal,mark=*,thick}},
    tick label style={font=\small},
    label style={font=\small\bfseries},
]

\end{axis}
\end{tikzpicture}
\caption{Max enclaves increase with PMP entries until exhaustion caps growth.}
\end{figure}
}
\end{frame}

\begin{frame}{PMP Misalignment: Why It Matters}
    \only<1>{
\begin{block}{The Problem}
PMP entries must describe power-of-two, naturally aligned regions (NAPOT). Misaligned memory allocations cause unnecessary padding.
\end{block}
    }
%     \only<2>{
% \begin{block}{The Effect}
% \begin{itemize}
%     \item Misalignment causes a single memory region to consume multiple PMP entries.
%     \item Fragmentation leads to rapid exhaustion of PMP resources (max = 16 entries).
% \end{itemize}
% \end{block}

>>>>

\begin{frame}{Enclave Overhead and System Behavior}
\small
% First overlay: text only
\only<1>{
\begin{block}{Performance Overhead}
\begin{itemize}
    \item \textbf{Enclave execution} introduces overhead:
        \begin{itemize}
            \item Compute-bound workloads like CoreMark and Dhrystone incur moderate overhead of 1.06$\times$ to 1.61$\times$.
            \item Hybrid workloads (Compute- and memory-bound) like Kyber suffer severe overhead, up to 2.73$\times$.
        \end{itemize}
    %\item \textbf{Memory-bound workloads} experience significant penalties due to enclave-induced memory isolation and security checks.
\end{itemize}
\end{block}

%\vspace{0.5cm} % Adjust space as needed

\centering
\begin{tikzpicture}
\begin{axis}[
    ybar,
    bar width=14pt,
    width=0.8\linewidth,
    height=4cm,
    ymin=0, ymax=120,
    ylabel={Rel. perf. (\%)},
    symbolic x coords={CoreMark,Dhrystone,Kyber},
    xtick=data,
    ymajorgrids,
    grid style=dashed,
    nodes near coords,
    every node near coord/.append style={font=\footnotesize, /pgf/number format/fixed},
    legend style={at={(0.5,-0.35)}, anchor=north, legend columns=-1}
]

% Native: orange
\addplot[
    fill={rgb,255:red,230; green,159; blue,0},
    fill opacity=0.8,
    draw=black!60,
    rounded corners=2pt
] coordinates {(CoreMark,100) (Dhrystone,100) (Kyber,100)};

% Enclave: blue
\addplot[
    fill={rgb,255:red,0; green,114; blue,178},
    fill opacity=0.8,
    draw=black!60,
    rounded corners=2pt
] coordinates {(CoreMark,95) (Dhrystone,62) (Kyber,37)};

}

% \only<2>{
% % \begin{frame}{Performance Trends and Hardware Bottlenecks}
% \centering
% \begin{figure}[htbp]
% \centering
% \begin{tikzpicture}
% \begin{axis}[
%     ybar,
%     bar width=14pt,
%     width=10cm,
%     height=6cm,
%     ymin=0, ymax=120,
%     ylabel={Relative performance (\%)},
%     symbolic x coords={CoreMark,Dhrystone,Kyber},
%     xtick=data,
%     ymajorgrids,
%     grid style=dashed,
%     nodes near coords,
%     every node near coord/.append style={font=\footnotesize, /pgf/number format/fixed},
%     legend style={at={(0.5,-0.15)}, anchor=north, legend columns=-1}
% ]

% % Native: orange, rounded top, tranRecommendationssparency
% \addplot[
%     fill={rgb,255:red,230; green,159; blue,0},
%     fill opacity=0.8,
%     draw=black!60,
%     rounded corners=2pt
% ] coordinates {(CoreMark,100) (Dhrystone,100) (Kyber,100)};

% % Enclave: blue, rounded top, transparency
% \addplot[
%     fill={rgb,255:red,0; green,114; blue,178},
%     fill opacity=0.8,
%     draw=black!60,
%     rounded corners=2pt
% ] coordinates {(CoreMark,95) (Dhrystone,62) (Kyber,37)};

% \legend{Native,Enclave}
% \end{axis}
% \end{tikzpicture}
% \caption{Relative performance of enclave vs native execution.}
% \end{figure}
% }
\end{frame}

% % Second overlay: add figure
% \only<3>{
% \begin{block}{Overhead Breakdown}
% \begin{itemize}
%     \item Memory isolation via PMP adds fixed latency per memory access.
%     \item Context switches between host and enclave domains are costly.
% \end{itemize}
% \end{block}
% }
% \centering
% \textbf{Key Insight:} Overhead varies based on workload characteristics and memory behavior.
% \end{frame}
% Figure appears only on the second slide of this frame

\begin{frame}{Impact of Scalability Limitations}
\small
\only<1>{
\begin{block}{CPU Scaling Limits}
\begin{itemize}
    \item Increasing CPU cores \textbf{benefits multi-threaded tasks}, but gains plateau due to context-switch overhead.
    \item For example, \texttt{CoreMark} shows marginal improvements beyond 4 cores.
\end{itemize}
\end{block}
\centering
\begin{tikzpicture}
\begin{axis}[
    ybar,
    bar width=8pt,
    width=0.75\linewidth,
    height=4.2cm,
    enlarge x limits=0.15,
    ymin=95, ymax=129,
    xlabel={CPU cores},
    ylabel={Rel.\ perf.\ (\%)},
    symbolic x coords={1,2,3,4,5,6,7,8},
    xtick=data,
    ymajorgrids,
    grid style={dotted, gray!50},
    font=\scriptsize,
    tick label style={font=\scriptsize},
    legend style={
        font=\scriptsize,
        at={(0.5,-0.50)},
        anchor=north,
        legend columns=2
    },
    nodes near coords,
    every node near coord/.append style={font=\tiny},
    point meta=explicit symbolic
]

@@

\begin{frame}{PMP Entry Exhaustion and Multi-Enclave Scalability}
\small
\only<1>{
\begin{block}{PMP Resource Limitations}
\begin{itemize}
    \item \textbf{Maximum PMP entries}: 16.
    \item Exhausted PMP entries \textbf{prevent new enclave creation}, causing errors even with abundant system resources.
    \item Failure due to PMP exhaustion: \texttt{SBI error codes}.
\end{itemize}
\end{block}
}
\only<2>{
\begin{block}{Scalability Bottleneck}
\begin{itemize}
    \item In \textbf{multi-tenant environments}, PMP exhaustion limits the number of concurrent enclaves.
    \item This creates scalability issues, especially for high-concurrency workloads.
\end{itemize}
\end{block}

\centering
\begin{tikzpicture}
\begin{axis}[
    width=10cm,
    height=4.5cm,
    xlabel={PMP Entries},
    ylabel={Concurrent Enclaves},
    xmin=0, xmax=16,
    ymin=0, ymax=16,
    xtick={0,4,8,12,16},
    ymajorgrids,
    grid style=dashed,
    mark options={solid},
    nodes near coords,
    every node near coord/.append style={font=\footnotesize, /pgf/number format/fixed},
    clip=false,
    cycle list={{teal,mark=*,thick}},
    tick label style={font=\small},
    label style={font=\small\bfseries},
]

\end{axis}
\end{tikzpicture}
}

@@

% \begin{frame}{PMP Entries vs Max Concurrent Enclaves}
% \centering
% \only<3>{
% \begin{figure}[htbp]
% \centering
% \begin{tikzpicture}
% \begin{axis}[
%     width=10cm,
%     height=6cm,
%     xlabel={PMP Entries},
%     ylabel={Concurrent Enclaves},
%     xmin=0, xmax=16,
%     ymin=0, ymax=16,
%     xtick={0,4,8,12,16},
%     ymajorgrids,
%     grid style=dashed,
%     mark options={solid},
%     nodes near coords,
%     every node near coord/.append style={font=\footnotesize, /pgf/number format/fixed},
%     clip=false,
%     cycle list={{teal,mark=*,thick}},
%     tick label style={font=\small},
%     label style={font=\small\bfseries},
% ]

% % Data plot
% \addplot[
%     color=teal,
%     mark=*,
%     thick
% ] coordinates {
%     (0,0) (2,2) (4,4) (6,6) (8,8) (10,8) (12,8) (14,8) (16,8)
% };

% % Highlight PMP exhaustion area
% \fill[red!30,opacity=0.4] (axis cs:8,0) rectangle (axis cs:16,16);

% % Add Yellow Sliding Zone (between 4 and 8 PMP entries)
% \fill[yellow!30, opacity=0.6] (axis cs:4,0) rectangle (axis cs:8,16);

% % Annotation inside shaded region
% \node[red, align=center, font=\footnotesize\bfseries] at (axis cs:12,12) {PMP Exhaustion \\ Caps at 8 Enclaves};
% \node[align=center, font=\footnotesize\bfseries, text=black] at (axis cs:6,12) {Sliding Zone};

% % Stability boundary line & label
% \draw[dashed, thick, color=blue] (axis cs:4,0) -- (axis cs:4,16);
% \node[anchor=south, font=\footnotesize\bfseries, color=blue] at (axis cs:4,16) {Stability Boundary};

% \end{axis}
% \end{tikzpicture}
% \caption{Max enclaves increase with PMP entries until exhaustion caps growth.}
% \end{figure}
% }
\end{frame}

\begin{frame}{PMP Misalignment: Why It Matters}
    \only<1>{
\begin{block}{The Problem}
PMP entries must describe power-of-two, naturally aligned regions (NAPOT). Misaligned memory allocations cause unnecessary padding.
\end{block}

\begin{tikzpicture}[scale=1, transform shape, font=\small]

% Memory bar
\draw[thick] (0,0) rectangle (12,1);
\node[anchor=south] at (0,1.1) {\textbf{Base}};
\node[anchor=south] at (12,1.1) {\textbf{Top}};

% Needed regions (non-contiguous)
\draw[fill=gray!30] (0.5,0.1) rectangle (2.2,0.9);
\node[font=\footnotesize] at (1.35,0.5) {\textbf{R1}};

\draw[fill=gray!30] (3.8,0.1) rectangle (4.6,0.9);
\node[font=\footnotesize] at (4.2,0.5) {\textbf{R2}};

\draw[fill=gray!30] (7.0,0.1) rectangle (9.1,0.9);
\node[font=\footnotesize] at (8.05,0.5) {\textbf{R3}};

% NAPOT allocations (outer boxes) and padding (hatched)
\draw[thick, fill=blue!20, draw=blue!80!black] (0.25,-0.3) rectangle (2.5,-1.1);
\node[font=\footnotesize, align=center] at (1.375,-0.7) {\textbf{NAPOT} \\ \footnotesize fit R1};

\draw[pattern=north east lines, pattern color=black] (0.25,-0.3) rectangle (0.5,-1.1);
\draw[pattern=north east lines, pattern color=black] (2.2,-0.3) rectangle (2.5,-1.1);

\draw[thick, fill=green!20, draw=green!60!black] (3.5,-0.3) rectangle (5.1,-1.1);
\node[font=\footnotesize, align=center] at (4.3,-0.7) {\textbf{NAPOT} \\ \footnotesize fit R2};

\draw[pattern=north east lines, pattern color=black] (3.5,-0.3) rectangle (3.8,-1.1);
\draw[pattern=north east lines, pattern color=black] (4.6,-0.3) rectangle (5.1,-1.1);

\draw[thick, fill=orange!20, draw=orange!70!black] (6.7,-0.3) rectangle (9.3,-1.1);
\node[font=\footnotesize, align=center] at (8.0,-0.7) {\textbf{NAPOT} \\ \footnotesize fit R3};

\draw[pattern=north east lines, pattern color=black] (6.7,-0.3) rectangle (7.0,-1.1);
\draw[pattern=north east lines, pattern color=black] (9.1,-0.3) rectangle (9.3,-1.1);

% Legend / note
\node[anchor=west] at (0,-1.6) {\footnotesize \textbf{Note:} Hatched = padding/waste due to power-of-two alignment};

    }
%     \only<2>{
% \begin{block}{The Effect}
% \begin{itemize}
%     \item Misalignment causes a single memory region to consume multiple PMP entries.
%     \item Fragmentation leads to rapid exhaustion of PMP resources (max = 16 entries).
% \end{itemize}
% \end{block}
